The Success of College Basketball Teams can be predicted using several advanced team statistics

### 2022-04-25 Paper-On-NCAAB-Stats/outputs/paper/Paper-On-NCAAB-Stats.Rmd
Final Results section

#Load in libraries
library(tidyverse)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggpubr)
library(modelsummary)
library(DiagrammeR)
library(tidymodels)
library(broom)
```

@@

```{r dags, echo=FALSE, fig.cap="A DAG showing Offensive Rating as a mediator"}
DiagrammeR::grViz("
digraph {
  graph [ranksep = 0.2]
  node [shape = plaintext, fontsize = 10, fontname = Helvetica]
    A [label = 'True Shooting % / Total Rebound % / Turnover %']
    B [label = 'Offensive Rating']
    C [label = 'Simple Rating System']
  edge [minlen = 2, arrowhead = vee]
    A -> C
    A -> B
    B->C
  { rank = same; A; C}
}
", height = 300)
```
In the DAG, it can be seen how true shooting percentage, total rebound percentage, and turnover percentage all affect and have causality with simple rating system. It can also be seen that offensive rating affects simple rating system too, but it is also affected by the three variables listed above and is being used as a mediator between these variables and simple rating system. As this mediator, offensive rating is explaining exactly why these three variables affect simple rating system [@citeBhandari]. The three variables affect team success because the three variables cause teams to either score more or less points on offense, giving them either a better or worse offensive rating, and thus causing them to either win or lose more games, which is essentially what the DAG is explaining. As a result of this, offensive rating should not be used in the model as the mediator will affect the strength of the relationship between the true predictor variables and simple rating system.

@@

```{r, echo = FALSE}
library(tidymodels)

@@

```{r, include=FALSE}
mods <- list(
  "mod_ts_trb_tov" = lm(srs ~ ts_percent + trb_percent + tov_percent, data = ncaab_stats_train),
  "mod_ts_trb" = lm(srs ~ ts_percent + trb_percent, data = ncaab_stats_train),
  "mod_ts_tov" = lm(srs ~ ts_percent + tov_percent, data = ncaab_stats_train),
  "mod_trb_tov" = lm(srs ~ trb_percent + tov_percent, data = ncaab_stats_train),
  "mod_ts" = lm(srs ~ ts_percent, data = ncaab_stats_train),
  "mod_trb" = lm(srs ~ trb_percent, data = ncaab_stats_train),
  "mod_tov" = lm(srs ~ tov_percent, data = ncaab_stats_train)
)
```

```{r gof, echo=FALSE}
modelsummary(mods,
             output = "kableExtra",
             statistic = "se = {std.error}",
             gof_omit = "Num.Obs.|Log.Lik.|F",
             fmt = 2,
             title = "Comparing goodness of fit measures of several potential models predicting Simple Rating System. Column names represent which variables are included in each model.")
```

@@

* _$\hat{y_i}$ represents the predicted value of SRS given specific values of the predictor variables_
* _$\hat{\beta_{0}}$ is the estimated intercept coefficient_
* _$x_1$ represents the predictor variable of True Shooting Percentage_
* _$\hat{\beta_{1}}$ is the estimated slope coefficient for True Shooting Percentage_
* _$x_2$ represents the predictor variable of Total Rebound Percentage_
* _$\hat{\beta_{2}}$ is the estimated slope coefficient for Total Rebound Percentage_
* _$x_3$ represents the predictor variable of Turnover Percentage_
* _$\hat{\beta_{3}}$ is the estimated slope coefficient for Turnover Percentage_
* _and $\epsilon$ is an error term_

@@

WRITE MODEL HERE MAYBE AND TALK ABOUT HOW YOU ADD VALUES TO PREDICT SRS

@@

@@

```{r}
library(plot3D)

scatter3D(ncaab_stats_train$ts_percent, ncaab_stats_train$trb_percent, ncaab_stats_train$srs, colvar = ncaab_stats_train$tov_percent, bty = 'b2',
          pch = 19, cex = 0.5, xlab = "TS%", ylab = "TRB%", zlab = "SRS", clab = "TOV%", ticktype = "detailed", )
```

@@

>>>>

#Load in libraries
library(tidyverse)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggpubr)
#install.packages("modelsummary")
library(modelsummary)
#install.packages("DiagrammeR")
library(DiagrammeR)
#install.packages("tidymodels")
library(tidymodels)
#install.packages("broom")
library(broom)
library(plot3D)
```

@@

```{r dags, echo=FALSE, fig.cap="A DAG showing Offensive Rating as a mediator"}
#Create Dag with arrow point from predictor variables to both simple rating system and offensive rating, then another arrow from offensive rating to SRS

DiagrammeR::grViz("
digraph {
  graph [ranksep = 0.2]
  node [shape = plaintext, fontsize = 10, fontname = Helvetica]
    A [label = 'True Shooting % / Total Rebound % / Turnover %']
    B [label = 'Offensive Rating']
    C [label = 'Simple Rating System']
  edge [minlen = 2, arrowhead = vee]
    A -> C
    A -> B
    B->C
  { rank = same; A; C}
}
", height = 300)
```
In the DAG, it can be seen how true shooting percentage, total rebound percentage, and turnover percentage all affect and have causality with simple rating system. It can also be seen that offensive rating affects simple rating system too, but it is also affected by the three variables listed above and is being used as a mediator between these variables and simple rating system. As this mediator, offensive rating is explaining exactly why these three variables affect simple rating system [@citeBhandari]. The three variables affect team success because the three variables cause teams to either score more or less points on offense, giving them either a better or worse offensive rating, and thus causing them to either win or lose more games, which is essentially what the DAG is explaining. As a result of this, offensive rating should not be used in the model as the mediator will affect the strength of the relationship between the true predictor variables and simple rating system.

@@

```{r, echo = FALSE}
#Split the dataset into a two different ones, one for training and one to test the model
#Split it 80:20

@@

```{r, include=FALSE}
#Create seven different models using our remaining three potential predictor variables
#Name each model based on which variables are included in it
#Assign all models to a single variable for later use

mods <- list(
  "mod_ts_trb_tov" = lm(srs ~ ts_percent + trb_percent + tov_percent, data = ncaab_stats_train),
  "mod_ts_trb" = lm(srs ~ ts_percent + trb_percent, data = ncaab_stats_train),
  "mod_ts_tov" = lm(srs ~ ts_percent + tov_percent, data = ncaab_stats_train),
  "mod_trb_tov" = lm(srs ~ trb_percent + tov_percent, data = ncaab_stats_train),
  "mod_ts" = lm(srs ~ ts_percent, data = ncaab_stats_train),
  "mod_trb" = lm(srs ~ trb_percent, data = ncaab_stats_train),
  "mod_tov" = lm(srs ~ tov_percent, data = ncaab_stats_train)
)
```

```{r gof, echo=FALSE}
#Create a summary table including parameter statistics and goodness of fit measures for each of the models
modelsummary(mods,
             output = "kableExtra",
             statistic = "se = {std.error}",
             gof_omit = "Num.Obs.|Log.Lik.|F",
             fmt = 2,
             title = "Comparing goodness of fit measures of several potential models predicting Simple Rating System. Column names represent which variables are included in each model.")
```

@@

* _$\hat{y_i}$ represents the predicted value of SRS given specific values of the predictor variables_
* _$\hat{\beta_{0}}$ is the estimated intercept coefficient_
* _$x_1$ represents the predictor variable of True Shooting Percentage_
* _$\hat{\beta_{1}}$ is the estimated slope coefficient for True Shooting Percentage_
* _$x_2$ represents the predictor variable of Total Rebound Percentage_
* _$\hat{\beta_{2}}$ is the estimated slope coefficient for Total Rebound Percentage_
* _$x_3$ represents the predictor variable of Turnover Percentage_
* _$\hat{\beta_{3}}$ is the estimated slope coefficient for Turnover Percentage_
* _and $\epsilon$ is an error term which shoud be 0 assuming our model is unbiased_

@@

@@

We can plug all these coefficients into the $\beta's$ in the formula explained in Section \@ref(model) to get the completed formula for our final model, $\hat{y_i} = -98.62 + 0.92x_1 + 1.63x_2 - 2.04x_3$. Doing so makes our model much easier to interpret. In order to estimate what a teams simple rating system value would be, you plug in their true shooting percentage for $x_1$, their total rebound percentage for $x_2$, and their turnover percentage for $x_3$, and the output of $\hat{y_i}$ would be their rating. For example, if a team has a true shooting percentage of 50%, a total rebounding percentage of 40%, and a turnover percentage of 20%, the model estimates that their simple rating system would be $\hat{y_i} = -98.62 + 0.92(50) + 1.63(40) - 2.04(20) = -28.22 $. This shows how the model can successfully predict a teams success with the use of advanced team statistics such as our predictor variables that were used, answering the research question.

@@

The relationship between simple rating system and our three explanatory variables can also be visualized despite being a 4D relationship, thanks to the package plot3D [@citePlot3D].

```{r 3d, echo=FALSE, fig.cap="4D plot showing the relationship between the models response variable, Simple Rating System, and its three predictor variables"}

scatter3D(ncaab_stats_train$ts_percent, ncaab_stats_train$trb_percent, ncaab_stats_train$srs, colvar = ncaab_stats_train$tov_percent, bty = 'b2',
          pch = 19, cex = 0.5, xlab = "TS%", ylab = "TRB%", zlab = "SRS", clab = "TOV%", ticktype = "detailed", )
```
Figure \@ref(fig:3d) shows this relationship through the use of a 3D plot, which is technically a 4D plot based on the use of a colour gradient to show a 4th variable. The vertical z-axis represents our response variable SRS, the x-axis closest to the viewer represents true shooting percentage, and the remaining y-axis on the right represents total rebound percentage. Also, the colour gradient is used to represent turnover percentage where each point is coloured in differently based on its turnover percentage value, giving the plot its 4th dimension.

Although it is difficult, it can be seen that the overall trend of the points moves from the bottom left of the plot towards the top right of the plot. So, as true shooting percentage increases along the x-axis and the points move right, simple rating system increases since the points are also moving upwards. Additionally, as total rebound percentage increases up the y-axis, simple rating system increases as well since the points are moving from the bottom of the z-axis to the top. Finally, the points towards the bottom of the plot where SRS is low are mostly red, orange and yellow. Then as SRS increases, the points gradually become more green and then more blue. This final dimension shows that as turnover percentage decreases and the colours go from red to blue, simple rating system increases. Each of these relationships shown by the 4D plot line up with what has been seen in both the graphical analysis and the analysis of the models parameter statistics. It should be noted that all three of these relationships are in the presence of the other predictor variables, and when you combine them all together you get the upwards trend and colour change from red to blue seen in the plot, which shows us the overall relationship estimated by the model.

Before going over what we have learned from our model, there are two things that must be done. First, we need to check to see if all of the assumptions that where made with respect to the model were all satisfied, or else there will be some serious limitations to our findings. Then, we need to use our testing dataset to validate our model and see if these results can be used to make predictions on other datasets.

@@

The first assumption is the linearity assumption where all explanatory variables must have a linear relationship with the response, and this was confirmed to be true in Section \@ref(data) through the use of scatter plots. The explanatory variables also should not have a strong linear relationship with each other, and this was also already considered and dealt with earlier through the use of DAGs in Section \@ref(model). The next three assumptions will be checked using the plots in Figure \@ref(fig:assumplots). The next assumptions is the independence errors assumptions, as errors should be independent of one another and show no correlation. The bottom three plots in Figure \@ref(fig:assumplots), which show the relationship between each predictor variable and the models residuals, prove that this assumption is satisfied. This is because the points are all randomly scattered around a horizontal line at 0 on the y-axis for each one of the three plots, and there are no separated groups of points or noticeable trends that would suggest any correlation in the residuals. The third assumption is the homoscedasticity of errors assumption which simply means the models residual errors have constant variance. The plot in the top right of the figure shows the residuals have constant variance since the points have an upwards trend along an invisible diagonal line, meaning that variance is mostly similar and constant across all fitted values. Finally, the last assumption states that the errors should be normally distributed. The histogram in the top left of Figure \@ref(fig:assumplots) is used to see if this assumption is satisfied, and it should show a normal distribution if it is. The histogram has a distribution very similar to the distribution of our response variable simple rating system, which we previously determined that despite being slightly right skewed, the distribution is very close to being a normal distribution centered around its mean. This seems to be the case here as well, and while it is not perfect, it is still very close to being so, meaning it should not cause many issues with our model, if any at all, and the assumption can be considered satisfied.

```{r, include=FALSE}
test_predictions <- predict(final_model, newdata = ncaab_stats_test)

prediction_error <- ncaab_stats_test$srs - test_predictions

rmse_predictions <- sqrt(mean(prediction_error^2))

Finally, our model must be validated using the testing dataset. One way to do this is by making predictions on the test dataset, using the values of the explanatory variables in the data. We then calculate the RMSE using the errors between the predicted values and actual values of the testing data, and compare it to the RMSE of the original model built with the training data []. After using the model to make predictions on the testing data, it was found that the predicted values had a RMSE of 7.74. This is very similar to the original models RMSE based on the training data, which was 7.29. Since these values are very similar to one another, it means that our model did a good job at predicting values using the testing dataset, and thus it should also successfully be able to make predictions on other datasets, like for example future college basketball seasons. This means that the model has successfully been validated using the testing dataset, and we can now analyze what exactly was learned from our model.
